---
title: "Introduction to R for data analysis"
author: Peter Carbonetto
output:
  beamer_presentation:
    template: beamer.tex
    keep_tex: false
---

```{r knitr-options, echo=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,fig.align = "center",
                      results = "hide", fig.show = "hide", message = FALSE,
		      warning = FALSE)
```

Aims of workshop
================

*Add workshop aims here.*

Our project: analyze 2016 & 2017 Divvy data
===========================================

*Describe the Divvy data analysis project and aims.*

It's your choice
================

Your may choose to...

+ Use R on your laptop.

+ Use RStudio on your laptop.

+ Use R or RStudio (Desktop or Server) on the RCC cluster.

+ Pair up with your neighbour.

+ Follow what I do on the projector.

**Note:** If you use the RCC cluster I'm assuming you know how to set
up an interactive computing session with appropriate amount of time
and memory, load R or RStudio, and display graphics (e.g., using
ThinLinc). *I can help with this.*

Software we will use today
==========================

1. R

2. R packages `data.table` and `ggplot2`.

3. RStudio (optional).

**Note:** I'm assuming you have already installed R and/or RStudio on
your laptop, or you are using the RCC cluster.

Outline of workshop
===================

*Add outline here.*

Outline of initial setup
========================

*Add outline of initial setup here.*

Initial setup
=============

+ WiFi

+ Power outlets

+ YubiKeys

+ Pace, questions (e.g., keyboard shortcuts).

Download or "clone" git repository
==================================

The first step is to download the workshop packet onto your computer
or onto the RCC cluster. Go to:

+ [**github.com/rcc-uchicago/R-intro-divvy-2**][github-repo]

To download, click green **"Clone or download"** button.

Or, if you have **git**, you can run this command:

```{bash download-repo, eval=FALSE}
git clone https://github.com/rcc-uchicago/
  R-intro-divvy-2.git
```

(Note the URL in the git command should not contain any spaces.)

What's in the workshop packet
=============================

```
R-intro-divvy-2
  /analysis # Scripts implementing data analyses.
  /code     # Additional R code used in analyses.
  /data     # Original ("raw") data.
  /docs     # Additional workshop materials.
  /output   # Processed data & results files.
```

Open the slides on your computer
================================

The PDF is useful for copying & pasting code from the slides.

+ The PDF is in the [**"docs"**](../docs) folder of the workshop packet.

+ You can also view the PDF by clicking the **"docs"** item in the file
  listing on the GitHub webpage.

Download the Divvy data
=======================

+ Disk space required: at least **2 GB**.

+ Download the 2016 & 2017 data files from here:

    + [**www.divvybikes.com/system-data**][divvy-data]

+ Download them to the [**"data"**](../data) folder.

+ You should have 4 ZIP files:

    ```
    Divvy_Trips_2016_Q1Q2.zip
    Divvy_Trips_2016_Q3Q4.zip
    Divvy_Trips_2017_Q1Q2.zip
    Divvy_Trips_2017_Q3Q4.zip
    ```

+ Unzip all of these files.

+ After unziping these files, you should have **15** CSV files.

[github-repo]: https://github.com/rcc-uchicago/R-intro-divvy-2
[divvy-data]: https://www.divvybikes.com/system-data

Set up your R environment
=========================

+ Launch R or RStudio.

+ *We will run all the code from the **"analysis"** folder*.

+ To change your working directory:

    + In R, use `getwd()` function.

    + In RStudio, select **Session > Set Working Directory > Choose
      Directory...`**

Before continuing, check your working directory:

```{r check-wd}
getwd()  # Should be .../analysis
```

Run `sessionInfo()`
===================

Check the version of R that you are using:

```{r check-version}
sessionInfo()
```

If you are using an older version of R (version 3.2 or earlier), you
should upgrade to the latest version. *Some of the examples may not
work in older versions of R.*

Check your R environment
========================

The R environment is where all variables and functions are stored and
accessed. Your R environment should be empty:

```{r check-env}
ls()
```

If you do see names of objects listed, it means your environment is
not empty, and you should restart R with a clean environment. (In
RStudio, go to **Session > Restart R**.)

Create a file to keep track of your analysis code
=================================================

+ In RStudio, select **File > R Script**.

+ Alternatively, use your favourite editor (emacs, vim, Atom, *etc*).

+ Add some comments to the tile to remind yourself what this file is
  for, e.g.,

    ```
    # R code I used to analyze Divvy data
	# during RCC workshop on May 1, 2018.
    ```

+ Save the file in the [**"analysis"**](../analysis) folder.
  Name the file whatever you'd like (e.g., `myanalysis.R`).

+ At this point, we are ready to start our analysis of the Divvy data.

Outline of workshop
===================

*Add outline of workshop here.*

Import the station data into R
==============================

Load the most recent station data into an R "data frame":

```{r read-station-data}
stations <-
  read.csv("../data/Divvy_Stations_2017_Q3Q4.csv",
           stringsAsFactors = FALSE)
```

This creates a new "data frame" object in your environment:

```{r inspect-station-data-1}
ls()
class(stations)
```

What does `read.csv` do, and what is a "data frame"? We can find the
answers by running

```{r help-readcsv-dataframe, eval=FALSE}
help(read.csv)
help(data.frame)
```

Inspect the Divvy station data
==============================

Run these commands to inspect the station data:

```{r inspect-station-data}
nrow(stations)
ncol(stations)
head(stations)
tail(stations)
stations[1:6,]
summary(stations)
```

*What do we learn about the station data from running these
commands? Did we reveal any issues with these data?*

Take a closer look at the "dpcapacity" column
=============================================

Run these commands to take a closer look at the "dpcapacity" column:

```{r inspect-dpcapacity-data-1}
x <- stations$dpcapacity
class(x)
summary(x)
table(x)
quantile(x,seq(0,1,0.1))
```

*What did we learn about the Divvy stations from running these
commands?*

Take an even closer look at dpcapacity data
===========================================

It is interesting that a few of the Divvy bike stations are much
larger than the others, and some have no docks. Where are these
stations?

```{r inspect-dpcapacity-data-2}
subset(stations,dpcapacity == 0)
subset(stations,dpcapacity >= 40)
```

Alternatively, we can sort the table rows and inspect the top and
bottom rows:

```{r inspect-dpcapcity-data-3}
rows <- order(stations$dpcapacity,decreasing=TRUE)
stations <- stations[rows,]
head(stations)
tail(stations)
```

Take a closer look at the "city" column
=======================================

Above we inspected *numeric* data. How can we inspect non-numeric data
in R?

```{r inspect-city-data-1}
x <- stations$city
class(x)
summary(x)
```

Unfortunately, the summary is not very useful here. The key is to
convert the "city" column to a "factor" (a categorical variable):

```{r inspect-city-data-2}
x <- factor(stations$city)
class(x)
summary(x)
```

*What did we learn about the Divvy stations from running these
commands? What issue did we discover in the station data? How could we
fix this issue?*

Import the Divvy trip data into R
=================================


