---
title: "Introduction to R for data analysis"
author: Peter Carbonetto
output:
  beamer_presentation:
    template: beamer.tex
    keep_tex: false
---

```{r knitr-options, echo=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,fig.align = "center",
                      results = "hide", fig.show = "hide", message = FALSE,
		      warning = FALSE)
```

Aims of workshop
================

*Add workshop aims here.*

Our project: analyze 2016 & 2017 Divvy data
===========================================

*Describe the Divvy data analysis project and aims.*

It's your choice
================

Your may choose to...

+ Use R on your laptop.

+ Use RStudio on your laptop.

+ Use R or RStudio (Desktop or Server) on the RCC cluster.

+ Pair up with your neighbour.

+ Follow what I do on the projector.

**Note:** If you use the RCC cluster I'm assuming you know how to set
up an interactive computing session with appropriate amount of time
and memory, load R or RStudio, and display graphics (e.g., using
ThinLinc). *I can help with this.*

Software we will use today
==========================

1. R

2. R packages `data.table` and `ggplot2`.

3. RStudio (optional).

**Note:** I'm assuming you have already installed R and/or RStudio on
your laptop, or you are using the RCC cluster.

Outline of workshop
===================

*Add outline here.*

Outline of initial setup
========================

*Add outline of initial setup here.*

Initial setup
=============

+ WiFi

+ Power outlets

+ YubiKeys

+ Pace, questions (e.g., keyboard shortcuts).

Download or "clone" git repository
==================================

The first step is to download the workshop packet onto your computer
or onto the RCC cluster. Go to:

+ [**github.com/rcc-uchicago/R-intro-divvy-2**][github-repo]

To download, click green **"Clone or download"** button.

Or, if you have **git**, you can run this command:

```{bash download-repo, eval=FALSE}
git clone https://github.com/rcc-uchicago/
  R-intro-divvy-2.git
```

(Note the URL in the git command should not contain any spaces.)

What's in the workshop packet
=============================

```
R-intro-divvy-2
  /analysis # Scripts implementing data analyses.
  /code     # Additional R code used in analyses.
  /data     # Original ("raw") data.
  /docs     # Additional workshop materials.
  /output   # Processed data & results files.
```

Open the slides on your computer
================================

The PDF is useful for copying & pasting code from the slides.

+ The PDF is in the [**"docs"**](../docs) folder of the workshop packet.

+ You can also view the PDF by clicking the **"docs"** item in the file
  listing on the GitHub webpage.

Download the Divvy data
=======================

+ Disk space required: at least **2 GB**.

+ Download the 2016 & 2017 data files from here:

    + [**www.divvybikes.com/system-data**][divvy-data]

+ Download them to the [**"data"**](../data) folder.

+ You should have 4 ZIP files:

    ```
    Divvy_Trips_2016_Q1Q2.zip
    Divvy_Trips_2016_Q3Q4.zip
    Divvy_Trips_2017_Q1Q2.zip
    Divvy_Trips_2017_Q3Q4.zip
    ```

+ Unzip all of these files.

+ After unziping these files, you should have **15** CSV files.

[github-repo]: https://github.com/rcc-uchicago/R-intro-divvy-2
[divvy-data]: https://www.divvybikes.com/system-data

Set up your R environment
=========================

+ Launch R or RStudio.

+ *We will run all the code from the **"analysis"** folder*.

+ To change your working directory:

    + In R, use `getwd()` function.

    + In RStudio, select **Session > Set Working Directory > Choose
      Directory...`**

Before continuing, check your working directory:

```{r check-wd}
getwd()  # Should be .../analysis
```

Run `sessionInfo()`
===================

Check the version of R that you are using:

```{r check-version}
sessionInfo()
```

If you are using an older version of R (version 3.2 or earlier), you
should upgrade to the latest version.

+ *Some of the examples may not work in older versions of R.*

Check environment
=================

The R environment is where all variables and functions are stored and
accessed. Your R environment should be empty:

```{r check-env}
ls()
```

If you do see names of objects listed, it means your environment is
not empty, and you should restart R with a clean environment. (In
RStudio, go to **Session > Restart R**.)

+ At this point, we are now ready to start our data analysis.

Outline of workshop
===================

*Add outline of workshop here.*

Take a peek at the Divvy station data
=====================================

Take a peek at the Divvy trip data
==================================


